---
title: "Introduction to progamming with R"
author: "B Reiss"
date: "2016-02-01"
output:
  ioslides_presentation:
    fig_caption: yes
    fig_retina: 1
    fig_width: 5
    fig_height: 4
    keep_md: yes
    smaller: no
    logo: logo_128.png
---

## Overview of the lab

- Background on programming 
- R concepts
- Exercise: Basic data analysis
- Help

## Programming (Wikipedia)

- Formulation of a problem that leads to a program

Tasks: 

- analysis, 
- developing understanding of the problem, 
- generating algorithm, 
- verification of algorithm, and 
- implementation (coding)  

Purpose: 

- find a sequence of instructions (algorithm) that will automate performing a specific task to solve a problem

## Remember

- Programming is something you do and not something you know. 

- It requires active practice. (Reading is not an active practice.)

## Programming structures

```{r, eval=FALSE}
Variables: 
-numeric, character, factor...

Control Structures: 
-if(), ifelse(), while(), for()...

Data Structures: 
- data.frame, class, matrix...

Syntax: 
-set of rules that defines the correct combinations of symbols
in that language

Tools:   
-library, functions...
```


## R

- R is a statistical programming language and a super advanced calculator

- > 2 million users

- based on functions that can be expanded through packages

- someone has already solved your problem: install.packages()

- it is an interpreter language each command is executed in the console

- "uses vectors to store data" 

## Help

help()

reference cards

books: 

- Beginning Data Science with R by Manas A. Pathak (free at UW)

- The Art of R Programming: A Tour of Statistical Software Design N. Matloff


## Function

- some INPUT - computation - some OUTPUT

- e.g. log10(): 10^3 = 1000 

- Thus log10(1000)=3

## Helper functions 

```{r, eval=FALSE}
help()
example()
c()
data.frame()
head();tail()
table()
summary()
str()
unique()
length()
subset()

assign:
  to <- from
  a <- 5
```

## Make your own function

```{r, eval=FALSE}
invmean<-function(data)
{
  
  rslt<-1/mean(data)
  return(rslt)
}

```

## Important functions

```{r, eval = FALSE}

apply()

sapply()

lapply()

by()

%in%

reshape()

```


## Workflow of a typical program

- get 
- review 
- clean 
- restructure
- summarize 
- analyse
- present results

## Getting data

 - External
 - Internal
 + Existing
 + New
 - Simulations

## External 

```{r, eval=FALSE}
read.csv() # won't work with a lot of EU data
read.tsv() 
read.table()
read.xlsx() # requires foreign package
```

## Internal

```{r, eval=FALSE}

# Existing
data()

# create
c()
seq()

#simulation - distributions
rnorm();rlnorm();runif()
```

## Variables

```{r, eval=FALSE}
#1-D
as.factor()
as.numeric()
as.character()


# mulit-D
c()
as.data.frame()
as.list()
as.matrix()

data.table() # -> through external package
```

## Special variable: time

```{r, eval=FALSE}

# time

as.POSIXct()
as.POSIXlt()

# packages: lubridate, chron

```

## Always use ISO time format

- 2015-01-12 23:17:12
- yyyy-mm-dd hh:mm:ss
- may include timezone

## Exercise

## Always start with a clean work space

```{r, eval=T}
rm(list=ls())
library(ggplot2)
library(doBy)
```

## Get data

```{r, eval=T}
data()

data(iris)

#current <- getwd()
#setwd(newdir)

dt <- iris

write.csv(dt,file = "irisdata.csv")
```

## Read the data back

```{r, eval=T}
read.csv(file = "irisdata.csv",sep = ",",nrows=5)

```

## Review

```{r, eval=T}
head(dt);tail(dt)
```

## Review

```{r, eval=T}
str(dt)
```

## Review

```{r, eval=T}
summary(dt)
```

## Review

```{r, eval=T}
unique(dt$Species)
```

## Review

```{r, eval=T}
table(dt$Species)
```

## Review

```{r, eval=T}
nrow(dt) 
```


## Create some data

```{r, eval=T}
dt$LLOD <- ifelse(dt$Sepal.Length < 5,"y","n")

str(dt)

table(dt$LLOD)
```

## Create contextual data: location, date, rain

```{r, eval=T}
location <- c("A","B","C")

date <- c("2015-11-01","2016-01-12")

rain <- c("lots","some","little","no","NA")

```


## Modify

```{r, eval=T}
dt$loc <- rep(location,150 / 3)

dt$date <- rep(sample(date),150 / 2)

dt$rain <- rep(rain,nrow(dt) / length(rain))

dt$rh <- runif(150,min = 37,max = 75)
```

## Modify

```{r, eval=T}
str(dt)
```

## Summarize

```{r, eval=T}
aggregate(Sepal.Length ~ loc + date, data = dt,FUN = "mean")
```

## Summarize

```{r, eval=T}
tbltst <-
  aggregate(cbind(Sepal.Length,rh) ~ loc + date + rain, 
            data = subset(dt,LLOD!="n") ,FUN = "mean")
head(tbltst)
```

## Summarize

```{r, eval=T}
tblsdt <-
  summaryBy(Sepal.Length ~ loc + date + rain,
            data = dt,FUN = c(mean,sd))
head(tblsdt)
```

## Summarize

```{r, eval=T}
colnames(tblsdt) <- c("lc","dt","rn","msl","sdsl")
str(tblsdt)
```

## Basic plots

```{r, eval=T}
boxplot(Sepal.Length ~ loc,data = dt)
dev.off()
```

## ggplot: The beauty of plotting

```{r, eval=T}
p <- ggplot(dt)

p <- p + geom_point(aes(x = Petal.Width ,y = Sepal.Length,color = loc))
```

## ggplot: The beauty of plotting

```{r, eval=TRUE}
p
dev.off()
```

## ggplot: The beauty of plotting

```{r, eval = TRUE}
p <- ggplot(dt,aes(x = Petal.Width , y = Sepal.Length))

p <- p + geom_point()

p <- p + geom_smooth(method = "lm", se = T)

p <- p + facet_grid(loc ~ date)
```

## ggplot: The beauty of plotting

```{r, eval = TRUE}
p
dev.off()
```

## Analysis

```{r, eval=T}
lm(Sepal.Length ~ loc + date + rain,data = tbltst)
```

## Analysis

```{r, eval=TRUE}
lf <- lm(Sepal.Length ~ loc + date + rain,data = tbltst)
```

## Analysis

```{r, eval=TRUE}
summary(lf)
```

## Presentation

RMarkDown



